---
title: "Using crumblr in practice"
subtitle: ''
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  toc_float: true
vignette: >
  %\VignetteIndexEntry{crumblr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---



<!---
cd /Users/gabrielhoffman/workspace/repos/crumblr/vignettes
R

rmarkdown::render('crumblr.Rmd')
--->

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(	tidy=FALSE, 
												cache=TRUE,
												echo=TRUE,
                      	dev=c("png", "pdf"),
                      	package.startup.message = FALSE,
                      	message=FALSE, 
                      	error=FALSE, 
                      	warning=FALSE
											  )

knitr::opts_chunk$set()

options(width=100)
```	



<style>
body {
text-align: justify}
</style>

Here we consider counts for 8 cell types from quantified using single cell RNA-seq data from unstimulated and interferon β stimulated PBMCs from 8 subjects [(Kang, et al., 2018)](https://www.nature.com/articles/nbt.4042).    

The functions here incorporate the precision weights:

- `variancePartition::fitExtractVarPartModel()` 
- `variancePartition::dream()`
- `limma::lmFit()`

## Differential testing
Here we evaluate whether the observed cell proportions change in response to interferon β.  Given the results here, we cannot reject the null hypothesis that interferon β does not affect the cell type proportions.
 
```{r crumblr}
library(crumblr)

# Load cell counts from Kang, et al. (2018)
#  https://doi.org/10.1038/nbt.4042
data(IFNCellCounts)

# Apply crumblr transformation 
# cobj is an EList object compatable with limma workflow
# cobj$E stores transformed values 
# cobj$weights stores precision weights
cobj = crumblr(cellCounts)

# Use variancePartition workflow to analyze each cell type
# Perform regression on each cell type separately
#  then use eBayes to shrink residual variance
# Also compatible with limma::lmFit() 
library(variancePartition)
fit = dream(cobj,  ~ StimStatus + ind, info)
fit = eBayes(fit)

# Extract results for each cell type
topTable(fit, coef="StimStatusstim", number=Inf, sort.by="none")
```

## PCA
Performing PCA on the transformed cell counts indicates that the samples cluster based on subject rather than stimulation status.

```{r pca, fig.height=5, fig.width=5}
library(ggplot2)

# Perform PCA
pca = prcomp(t(cobj$E))

# merge with metadata 
df_pca = merge(pca$x, info, by="row.names")

# Plot PCA 
#   color by Subject
# 	shape by Stimulated vs unstimulated
ggplot(df_pca, aes(PC1, PC2, color=as.character(ind),  shape=StimStatus)) + 
		geom_point(size=3) + 
		theme_classic() + 
		theme(aspect.ratio=1) + 
		scale_color_discrete(name="Subject") +
		xlab("PC1") + ylab("PC2")
```

## Hierachical clustering
The samples from the same subject also cluster together.
```{r hclust}
heatmap(cobj$E)
```

## Variance partitioning
Decomposing the variance illustrates that more variation is explained by subject than stimulation status.
```{r vp, fig.height=4, fig.width=6}
library(variancePartition)

# Partition variance into components for Subject (i.e. ind)
#   and stimulation status, and residual variation
vp = fitExtractVarPartModel( cobj, ~ ind + StimStatus, info)

# Plot variance fractions
plotPercentBars(vp)
```








